% !TEX root = document.tex

\chapter{\label{chap:Preface}Preface}
\addcontentsline{toc}{subsection}{Preface}

Almost three years. That is how long it took me to finish this thesis. During this period:

\begin{itemize}
 \item Startup companies were joined and left;
 \item A full-time employment was started;
 \item Many hobby projects were started;
 \item Not a single hobby project was finished.
\end{itemize}

Although anecdotal, the list above reflects my tendency to distract myself with somewhat useful things when I doubt my ability to finish the main objective.
Regardless of the time taken, I am happy with the result in the form of this thesis, which would not have been the same without the help and support of a lot of people.
I am sure I will forget to mention some people, but I will attempt to list most of them.

First and foremost, I want to thank my supervisor Danny Groenewegen for his patience, for sharing his experience and for his practical help.
In the same vein, I want to thank Jasper Cockx and Arie van Deursen who were able to give me the final push needed to finish this thesis.
Next, I want to thank my colleagues Elmer and Daniël, as well as all other members of the Programming Languages group for sharing their experience and motivating me to keep going.
Special thanks go to my parents, brother, and girlfriend for their unconditional support and trust.

Finally, this thesis would not exist without Eelco Visser.
Although he is no longer with us, his ideas, contagious positivity, and continued trust in me are the foundations on which this thesis is built, and I am grateful to have experienced his mentorship.

\vspace{1cm}
\begin{flushright}
\theauthor{}\\
Delft, the Netherlands\\
\today{}\\
\end{flushright}
